Geometrization of characters of tori over non-Archimedean local fields

### 2013-08-11 shanghai.tex
STILL working on the finite geochars lemmata.

We will see that the full subcategory $\GC_0(G)$ of bounded geometrized characters is a proper subcategory, in general. Indeed, it follows from our main result regarding geometrized characters, Theorem~\ref{thm:snake}, that the restriction of $\trFrob{G} : \GCiso{G} \to \Hom_\text{grp}(G(\Fq),\EEx)$ to the subgroup $\GCbiso{G}$ is an isomorphism onto the subgroup of $\chi \in \Hom_\text{grp}(G(\Fq),\EEx)$ such that the image of $\chi$ is bounded in $\EEx$.

@@

To define the category of finite geometrized characters we must work a bit harder.
Consider the following category \cdef{$C(G)$}:
an object in $C(G)$ is a pair $(f,\psi)$ where: 
$f : H\to G$ an isogeny (a finite, surjective etale morphism and a morphism of commutative group $\Fq$-schemes)
for which the action of $\Gal(\bFq/\Fq)$ on the etale group scheme $\ker f$ is trivial, 
and $\psi : \ker f\to \Aut(V)$ is a representation on a finite-dimensional $\EE$-vector space $V$;
a map $(f,\psi) \to (f',\psi')$ in $C(G)$ is a pair $(\alpha,T)$ 
where $\alpha$ is a morphism of schemes such that $f' = f\circ \alpha$
and $T : V\to V'$ is a linear transformation equivariant 
for the action of $\ker f'$ on $V'$ by $\psi'$ 
and the action of $\ker '$ on $V$ by $\psi \circ \alpha$.
\[
\begin{tikzcd}
V \arrow{d}{T} & \Aut(V) & \arrow[swap]{l}{\psi} \ker f \arrow[hook]{r} & H \arrow{r}{f} & \arrow[equal]{d} G\\
V' & \Aut(V) & \arrow[swap]{l}{\psi} \arrow[swap]{u}{\alpha\vert_{\ker f}} \ker f \arrow[hook]{r} & \arrow[swap]{u}{\alpha} H' \arrow{r}{f'} & G\\
\end{tikzcd}
\]
We say that $(\alpha,T)$ in $C(G)$ is a \cdef{weak isomorphism} if $\alpha$ is a covering and $T$ is an isomorphism of vector spaces.
\Clifton{I suspect $\alpha$ is a covering, for free. Is that right? 
If $f' = f \circ \alpha$ and $f$ and $f'$ are etale covers, 
does it follow that $\alpha$ is also an etale cover? (Etale: Yes. Finite? Surjective?) 
If so, then I can simplify the definition of weak isomorphism accordingly.}

We put a tensor category structure on $C(G)$ as follows.
Let $(f,\psi)$ and $(f',\psi')$ be objects in $C(G)$. 
The category of isogenies to $G$ admits products: 
write $f\times_G f' : H\times_G H' \to G$ for the product of $f : H\to G$ and $f' : H'\to G$. 
Since $\ker(f\times_G f') \iso (\ker f)\times (\ker f')$, canonically,
the tensor product of the representations $\psi : \ker f\to \Aut(V)$ and $\psi' : \ker f'\to \Aut(V')$ 
is the representation $\psi \otimes\psi' :  \ker(f\times_G f') \to \Aut(V\otimes V')$. 
This allows us to set $(f,\psi)\otimes(f',\psi') \ceq (f\times_G f' ,\psi\otimes\psi')$. 
The same fact makes sense of $(f,\psi)\oplus(f',\psi')\ceq (f\times_G f' ,\psi\oplus\psi')$. 
The dual of $(f,\psi)$ is $(f,\psi)^\vee \ceq (f,\psi^\vee)$ 
where $\psi^\vee : \ker f \to \Aut(V^\vee)$ and $V^\vee = \Hom_\text{vec}(V,\EE)$. 
The verification that these operations make $C(G)$ a tensor category is routine.

%
Let \cdef{$C(G)[W^{-1}]$} be the localization of $C(G)$ at weak isomorphisms.
Now, let \cdef{$C_1(G)$} be the subcategory of $C(G)$ such that $\psi$ is a one-dimensional representation
and let \cdef{$\GCf(G)$} be the localisation of $C_1(G)$ at weak isomorphisms. 
This is the rigid symmetric monoidal category of \cdef{finite geometrized characters} of $G$.

\begin{proposition}\label{prop:GCf-pull-back}
If $g : G'\to G$ is a morphism of group schemes then $(f,\psi) \mapsto (f_g,\psi_g)$ defines a monomial functor $g^* : \GCf(G) \to \GCf(G')$ where $f_g : H\times_G G'\to G'$ and $\psi_g : \ker f_g \to \Aut(V)$ are defined by the diagram below.
\[
\begin{tikzcd}
V & \Aut(V) & \arrow[swap]{l}{\psi} \ker f \arrow[hook]{r} & H \arrow{r}{f} &  G\\
  &  & \arrow[dashed]{ul}{\psi_g} \arrow[swap]{u}{(g_f)\vert_{\ker g_f}} \ker f_g \arrow[hook]{r} & \arrow[swap]{u}{g_f} H\times_G G' \arrow{r}{f_g} & \arrow{u}{g} G'.
\end{tikzcd}
\]
\end{proposition}

\begin{proposition}
Each $(f,\psi)\in \GCf(G)$ determines a canonical isomorphism $m^*(f,\psi) \to (f,\psi)\boxtimes(f,\psi)$ in $\GCf(G\times G)$.
 \end{proposition}

Every morphism of group schemes $g: G'\to G$ 
defines a function $g^*: \obj C(G) \to \obj C(G')$ 
by $g^*(f,\psi) = (f_g,\psi_g)$ where
$f_g : H\times_G G'\to G'$ and $\psi_g : \ker f_g \to \Aut(V)$ are defined by the diagram below.
\[
\begin{tikzcd}
V & \Aut(V) & \arrow[swap]{l}{\psi} \ker f \arrow[hook]{r} & H \arrow{r}{f} &  G\\
  &  & \arrow[dashed]{ul}{\psi_g} \arrow[swap]{u}{(g_f)\vert_{\ker g_f}} \ker f_g \arrow[hook]{r} & \arrow[swap]{u}{g_f} H\times_G G' \arrow{r}{f_g} & \arrow{u}{g} G'.
\end{tikzcd}
\]
This definition is

@@

 Now consider the localization $C(G)[W^{-1}]$ of $C(G)$ at weak isomorphisms.
 Morphisms in $C(G)[W^{-1}]$ are simple to describe because category $C(G)$ 
 admits a calculus of left fractions\footnote{See {\tt http://ncatlab.org/nlab/show/calculus+of+fractions}.} for weak isomorphisms: 
 every isomorphism in $C(G)$ is a weak isomorphism; 
 push-outs exist in $C(G)$ and the push-out of a weak isomorphism is a weak isomorphism; 
 and weak isomorphisms enjoy the left-cancellation property. 
 It follows that a morphism in $C(G)[W^{-1}]$ is the equivalence class of 
  \[
   \begin{tikzcd}
   (f,\psi) \arrow{r}{(\alpha_1,T_1)} & (f_1,\psi_1) & \arrow[swap]{l}{(\alpha'_1,T'_1)} (f',\psi')
   \end{tikzcd}
  \]
 where $(\alpha',T') : (f,\psi) \to (f',T')$ is any map in $C(G)$ 
 and $(\alpha'_1,T'_1) : (f_1,\psi_1) \to (f',\psi')$ is a weak isomorphism.
 \Clifton{Reveal the commented-out details . . . I think I need them!}
% this morphism is commonly denoted by $(\alpha_1,T_1)^{-1}\circ (\alpha'_1,T'_1)$. 
 %The equivalence relation identifies $(\alpha_1,T_1)^{-1}\circ (\alpha'_1,T'_1)$ 
 % with $(\alpha_2,T_2)^{-1}\circ (\alpha'_2,T'_2)$ if there are maps 
 %$(f_1,\psi_1)\to (f_0,\psi_0)$ and $(f_2,\psi_2)\to (f_0,\psi_0)$ making the following diagram commute
 %\[
 %  \begin{tikzcd}
 %  & (f_1,\psi_1) \arrow{d} & \\
 %  (f,\psi) \arrow{ur}{(\alpha_1,T_1)} \arrow{dr}{(\alpha_2,T_2)} & (f_0,\psi_0)  & \arrow[swap]{ld}{(\alpha'_2,T'_2)} \arrow[swap]{lu}{(\alpha'_1,T'_1)} (f',\psi')\\
 %  & (f_2,\psi_2) \arrow{u} & \\
 %  \end{tikzcd}
 %\]
 %where $(f',\psi') \to (f_0,\psi_0)$ is a weak isomorphism.

@@

 If $(f,\psi)$ in an object in $C_1(G)$ then, as we have already seen, 
 $m^*L(f,\psi) \to L(f,\psi) \boxtimes L(f,\psi)$ by a weak isomorphism in $C_1(G\times G)$.
 This morphism in $C_1(G\times G)$ is not unique, but its image in $\GCf(G\times G)$ is . . .!

 let $\mu(f,\psi)$ be that isomorphism of $\ell$-adic sheaves on $G$,
 It follows immediately that the functor $L_G  \ceq  L_{W}\vert_{\GCf(G)} : \GCf(G) \to \GCb(G)$, 
 given on objects by $(f,\psi) \mapsto (L(f,\psi), \mu(f,\psi))$,
 is a full and faithfull monoidal functor, completing the proof.
\end{proof}

>>>>

We will see that the full subcategory $\GC_0(G)$ of bounded geometrized characters is a proper subcategory of $\GC(G)$, in general. Indeed, it follows from our main result regarding geometrized characters, Theorem~\ref{thm:snake}, that the restriction of $\trFrob{G} : \GCiso{G} \to G(\Fq)^*$ to the subgroup $\GCbiso{G}$ is an isomorphism onto the subgroup of $\chi \in G(\Fq)^*$ such that the image of $\chi$ is bounded in $\EEx$, so if $G(\Fq)$ is not finite, then $\GC_0(G) \hookrightarrow\GC(G)$ is not essentially surjective.

@@

To define the category of finite geometrized characters we must work a bit harder.
Consider the following category \cdef{$C(G)$}:
an object in $C(G)$ is a pair $(f,\psi)$ where: 
$f : H\to G$ an isogeny (a finite, surjective etale morphism and a morphism of commutative group $\Fq$-schemes)
for which the action of $\Gal(\bFq/\Fq)$ on the etale group scheme $\ker f$ is trivial, 
and $\psi : \ker f\to \Aut(V)$ is a representation on a finite-dimensional $\EE$-vector space $V$;
a map $(f,\psi) \to (f',\psi')$ in $C(G)$ is a pair $(\alpha,T)$ 
where $\alpha$ is a morphism of schemes such that $f' = f\circ \alpha$
and $T : V\to V'$ is a linear transformation equivariant 
for the action of $\ker f'$ on $V'$ by $\psi'$ 
and the action of $\ker '$ on $V$ by $\psi \circ \alpha$.
\[
\begin{tikzcd}
V \arrow{d}{T} & \Aut(V) & \arrow[swap]{l}{\psi} \ker f \arrow[hook]{r} & H \arrow{r}{f} & \arrow[equal]{d} G\\
V' & \Aut(V) & \arrow[swap]{l}{\psi} \arrow[swap]{u}{\alpha\vert_{\ker f}} \ker f \arrow[hook]{r} & \arrow[swap]{u}{\alpha} H' \arrow{r}{f'} & G\\
\end{tikzcd}
\]
\Clifton{I suspect $\alpha$ is a covering, for free. Is that right? 
If $f' = f \circ \alpha$ and $f$ and $f'$ are etale covers, 
does it follow that $\alpha$ is also an etale cover? (Etale: Yes. Finite? Surjective?) 
If so, then I can simplify the definition of weak isomorphism accordingly.}
We say that $(\alpha,T)$ in $C(G)$ is a \cdef{weak isomorphism} if $\alpha$ is a covering and $T$ is an isomorphism of vector spaces.
Let \cdef{$C(G)[W^{-1}]$} be the localization of $C(G)$ at weak isomorphisms.

We put a tensor category structure on $C(G)$ 
and a rigid monoidal category structure on $C_1(G)$ as follows.
Let $(f,\psi)$ and $(f',\psi')$ be objects in $C(G)$. 
The category of isogenies to $G$ admits products: 
write $f\times_G f' : H\times_G H' \to G$ for the product of $f : H\to G$ and $f' : H'\to G$. 
Since $\ker(f\times_G f') \iso (\ker f)\times (\ker f')$, canonically,
the tensor product of the representations $\psi : \ker f\to \Aut(V)$ and $\psi' : \ker f'\to \Aut(V')$ 
is the representation $\psi \otimes\psi' :  \ker(f\times_G f') \to \Aut(V\otimes V')$. 
This allows us to set $(f,\psi)\otimes(f',\psi') \ceq (f\times_G f' ,\psi\otimes\psi')$. 
The same fact makes sense of $(f,\psi)\oplus(f',\psi')\ceq (f\times_G f' ,\psi\oplus\psi')$. 
The dual of $(f,\psi)$ is $(f,\psi)^\vee \ceq (f,\psi^\vee)$ 
where $\psi^\vee : \ker f \to \Aut(V^\vee)$ and $V^\vee = \Hom_\text{vec}(V,\EE)$. 
The verification that these operations make $C(G)$ a tensor category is straightforward.
Likewise, $C(G)[W^{-1}]$ is a tensor category.

Let \cdef{$C_1(G)$} be the rigid monoidal full subcategory of $C(G)$ consisting of objects $(f,\psi)$ such that $\psi$ is a one-dimensional representation and let \cdef{$\GCf(G)$} be the localisation of $C_1(G)$ at weak isomorphisms; this is the rigid symmetric monoidal category of \cdef{finite geometrized characters} of $G$.

@@

 Now consider the localization $C(G)[W^{-1}]$ of $C(G)$ at weak isomorphisms.
 Morphisms in $C(G)[W^{-1}]$ are simple to describe because category $C(G)$ 
 admits a calculus of left fractions\footnote{See {\tt http://ncatlab.org/nlab/show/calculus+of+fractions}.} for weak isomorphisms: 
 every isomorphism in $C(G)$ is a weak isomorphism; 
 push-outs exist in $C(G)$ and the push-out of a weak isomorphism is a weak isomorphism; 
 and weak isomorphisms enjoy the left-cancellation property. 
 It follows that a morphism in $C(G)[W^{-1}]$ is the equivalence class of 
  \begin{equation}\label{eq:morGCf}
   \begin{tikzcd}
   (f,\psi) \arrow{r}{(\alpha_1,T_1)} & (f_1,\psi_1) & \arrow[swap]{l}{(\alpha'_1,T'_1)} (f',\psi')
   \end{tikzcd}
  \end{equation}
 where $(\alpha',T') : (f,\psi) \to (f',T')$ is any map in $C(G)$ 
 and $(\alpha'_1,T'_1) : (f_1,\psi_1) \to (f',\psi')$ is a weak isomorphism,
 and therefore given by a diagram of the form

@@

 By Lemma~\ref{lemma:finite-iso}, each $(f,\psi) \in \GCf(G)$ determines 
 an isomorphism $\mu(f,\psi) : m^*(f,\psi) \to (f,\psi) \boxtimes (f,\psi)$ in $\GCf(G\times G)$, canonically.
 It follows immediately that the functor 
 \[
 L_G  \ceq  L_{W}\vert_{\GCf(G)} : \GCf(G) \to \GCb(G),
 \] 
 given on objects by $(f,\psi) \mapsto (L(f,\psi), L_{W}(\mu(f,\psi)))$,
 is a full and faithfull monoidal functor, completing the proof.
\end{proof}

\begin{lemma}\label{lemma:finite-pull-back}
If $g : G'\to G$ is a morphism of group schemes then $(f,\psi) \mapsto (f_g,\psi_g)$ defines a monomial functor $g^* : C(G)[W^{-1}] \to C(G')[W'^{-1}]$ where $f_g : H\times_G G'\to G'$ is a pull-back of $f$ along $g$ and $\psi_g : \ker f_g \to \Aut(V)$ is defined by the diagram below.
\[
\begin{tikzcd}
V & \Aut(V) & \arrow[swap]{l}{\psi} \ker f \arrow[hook]{r} & H \arrow{r}{f} &  G\\
  &  & \arrow[dashed]{ul}{\psi_g} \arrow[swap]{u}{(g_f)\vert_{\ker g_f}} \ker f_g \arrow[hook]{r} & \arrow[swap]{u}{g_f} H\times_G G' \arrow{r}{f_g} & \arrow{u}{g} G'.
\end{tikzcd}
\]
\end{lemma}

\begin{proof}
The main issue to address is well-definedness. Although the pull-back appearing in the diagram above it not unique, the recipe for $g^*$ is well-defined as a functor between localized categories.
To see this, recall that a morphism in $\GCf(G)$ is an equivalence class of a diagram like \eqref{eq:morGCf}, commonly denoted by $(\alpha_1,T_1)^{-1}\circ (\alpha'_1,T'_1)$. 
The equivalence relation identifies $(\alpha_1,T_1)^{-1}\circ (\alpha'_1,T'_1)$
 with $(\alpha_2,T_2)^{-1}\circ (\alpha'_2,T'_2)$ if there are maps 
 $(f_1,\psi_1)\to (f_0,\psi_0)$ and $(f_2,\psi_2)\to (f_0,\psi_0)$ making the following diagram commute
 \[
  \begin{tikzcd}
  \  & A & \\
B   & C & D \\
  & E &
  \end{tikzcd}
 \]
 where $(f',\psi') \to (f_0,\psi_0)$ is a weak isomorphism.

\end{proof}

\begin{lemma}\label{lemma:finite-iso}
Each $(f,\psi)\in \GCf(G)$ determines a canonical isomorphism $m^*(f,\psi) \to (f,\psi)\boxtimes(f,\psi)$ in $\GCf(G\times G)$.
 \end{lemma}
